---
title: "Using mitch package for Infinium methylation analysis of phyllodes methylation"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here we are comparing non-malignant vs malignant phyllodes samples.

Data provided by Prof Ruth Pidsley's team.

Source code: https://github.com/markziemann/phyllodes_pathways

Gene set information provided by Reactome, Gene Ontology, and transcription factor
targets from MsigDB.

## Requirements

Load packages.

Important: ensure that the mitch version used is 1.15.0 (patched), 1.15.1 or higher.

# devtools::install_github("markziemann/mitch") #get devel mitch

```{r,packages}

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("HGNChelper")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
  library("missMethyl")
  library("gridExtra")
  library("png")
})

```

## Load methylation data

The limma results are read in.
Results from Meyer et al (PMID:38300122).

```{r,loaddata}

dm <- read.csv("Phyllodes_topTable.csv.gz",header=TRUE,row.names=1)

dim(dm)

head(dm)

```

Note that this object has probes with their own column, not simply as row names.

## Load pathways

```{r,loadpathways}

reactome <- gmt_import("ReactomePathways_2024-03-10.gmt")

gobp <- gmt_import("c5.go.v2023.2.Hs.symbols.gmt")
gobp <- gobp[grep("GOBP_",names(gobp))]
names(gobp) <- gsub("_"," ",gsub("GOBP_","",names(gobp)))

tft <- gmt_import("c3.tft.gtrd.v2023.2.Hs.symbols.gmt")
names(tft) <- gsub("_"," ",names(tft))

```

## Curate the annotation

Curate the table which matches probes to gene names.

It is important to update defunct gene symbols.

```{r,anno1}

tic()
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt <- stack(gp2)
colnames(gt) <- c("gene","probe")
gt$probe <- as.character(gt$probe)
dim(gt)
str(gt)
toc() #9.0s

tic()
#new.hgnc.table <- getCurrentHumanMap()
new.hgnc.table <- readRDS("new.hgnc.table.rds")
fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
length(unique(fix2$x))
gt$gene <- fix$Suggested.Symbol
toc()

head(gt)
str(gt)

```

## Mitch pipeline

The first part is to import the data into mitch.

```{r,mg1}

m <- mitch_import(x=dm,DEtype="limma",geneTable=gt)

head(m) %>%
  kbl(caption = "Differential gene methylation scores used for pathway analysis") %>%
  kable_paper("hover", full_width = F)

```

Now run the enrichment analysis.

Note that the results are not sorted by p-value, rather S.distance, an enrichment score.
I think this works better for interpretation.

### Reactome

Reactome is the first analysis.

The S distance is the enrichment score, which 0 means no chance and has a maximum of +1 and minimum
of -1.

```{r,reactome1}

mres1 <- mitch_calc(x=m,genesets=reactome,minsetsize=5, priority="effect",cores=8)

mtable1 <- mres1$enrichment_result
up <- subset(mtable1,s.dist>0 & p.adjustANOVA<0.05)
dn <- subset(mtable1,s.dist<0 & p.adjustANOVA<0.05)
nrow(up)
nrow(dn)

head(up,10) %>%
  kbl(caption = "Top significant pathways with higher methylation") %>%
  kable_paper("hover", full_width = F)

head(dn,10) %>%
  kbl(caption = "Top significant pathways with lower methylation") %>%
  kable_paper("hover", full_width = F)

```

Now make a barplot of these top findings.

```{r, reactome_barplot}

top <- rbind(up[1:10,],dn[1:10,])
top <- top[order(top$s.dist),]

barnames <- gsub("_"," ",top$set)
cols <- as.character(((sign(top$s.dist)+1)/2)+1)
cols <- gsub("1","blue",gsub("2","red",cols))

par(mar = c(5.1, 29.1, 4.1, 2.1))
barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Reactomes")
grid()

par( mar = c(5.1, 4.1, 4.1, 2.1) )

```

### GO Biological Process

GOBP is a larger annotation set.

It looks like defensins are increased in methylation, while some lipid processes are reduced.

```{r,gobp1}

mres2 <- mitch_calc(x=m,genesets=gobp,minsetsize=5, priority="effect",cores=8)

mtable2 <- mres2$enrichment_result
up <- subset(mtable2,s.dist>0 & p.adjustANOVA<0.05)
dn <- subset(mtable2,s.dist<0 & p.adjustANOVA<0.05)
nrow(up)
nrow(dn)

head(up,10) %>%
  kbl(caption = "Top significant GO BPs with higher methylation") %>%
  kable_paper("hover", full_width = F)

head(dn,10) %>%
  kbl(caption = "Top significant GO BPs with lower methylation") %>%
  kable_paper("hover", full_width = F)

```

These results are more interesting than the reactome ones.

Now make a barplot of these top findings.

```{r, gobp_barplot}

top <- rbind(up[1:10,],dn[1:10,])
top <- top[order(top$s.dist),]

barnames <- gsub("_"," ",top$set)
cols <- as.character(((sign(top$s.dist)+1)/2)+1)
cols <- gsub("1","blue",gsub("2","red",cols))

par(mar = c(5.1, 29.1, 4.1, 2.1))
barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
  cex.axis=0.8,col=cols, xlab="Enrichment score",main="GO Biological Processes")
grid()

par( mar = c(5.1, 4.1, 4.1, 2.1) )

```

### Transcription factor targets

Might give some clues about potential mechanisms.

RBMX and NEUROD are interesting, but they are relatively small sets, however NFKBIA is more interesting.

```{r,tft1}

mres3 <- mitch_calc(x=m,genesets=tft,minsetsize=5, priority="effect",cores=8)

mtable3 <- mres3$enrichment_result
up <- subset(mtable3,s.dist>0 & p.adjustANOVA<0.05)
dn <- subset(mtable3,s.dist<0 & p.adjustANOVA<0.05)
nrow(up)
nrow(dn)

head(up,10) %>%
  kbl(caption = "Top significant transcription factor target sets with higher methylation") %>%
  kable_paper("hover", full_width = F)

head(dn,10) %>%
  kbl(caption = "Top significant transcription factor target sets with lower methylation") %>%
  kable_paper("hover", full_width = F)

```

Now make a barplot of these top findings.

```{r, tft_barplot}

top <- rbind(up[1:10,],dn[1:10,])
top <- top[order(top$s.dist),]

barnames <- gsub("_"," ",top$set)
cols <- as.character(((sign(top$s.dist)+1)/2)+1)
cols <- gsub("1","blue",gsub("2","red",cols))

par(mar = c(5.1, 29.1, 4.1, 2.1))
barplot(abs(top$s.dist),horiz=TRUE,las=1,names.arg=barnames, cex.names=0.8,
  cex.axis=0.8,col=cols, xlab="Enrichment score",main="Transcription factor target sets")
grid()

par( mar = c(5.1, 4.1, 4.1, 2.1) )

```

## Output reports and charts

Make a html report and some charts.

```{r,reports}

mitch_report(res=mres1,outfile="reactome_mitchreport.html",overwrite=TRUE)
mitch_plots(res=mres1,outfile="reactome_mitchcharts.pdf")

mitch_report(res=mres2,outfile="gobp_mitchreport.html",overwrite=TRUE)
mitch_plots(res=mres2,outfile="gobp_mitchcharts.pdf")

mitch_report(res=mres3,outfile="tft_mitchreport.html",overwrite=TRUE)
mitch_plots(res=mres3,outfile="tft_mitchcharts.pdf")

```

## Session information

```{r,save}

sessionInfo()

```
